
---
title: "Exploring School Engagement and Expectations Using LLM-Assisted Wrangling"
output:
  html_document:
    df_print: paged
---

## Introduction

This project explores how school climate benchmarks (such as engagement, safety, communication, and expectations) correlate with academic performance across NYC school districts. By integrating traditional data analysis with LLM-assisted insights, we aim to uncover how these benchmarks may help explain grade outcomes and district performance differences.

---

## Load Libraries and Dataset

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggcorrplot)
library(pheatmap)

# Load the dataset
school_data <- read_csv("C:/Users/efthi/Downloads/cleaned_school_data.csv")
head(school_data)
grade_scale <- c("A" = 4, "B" = 3, "C" = 2, "D" = 1, "F" = 0)
school_data$numeric_grade <- grade_scale[school_data$overallgrade]

```

---

## Correlation Heatmap of Numeric Features

```{r}
# Ensure all selected columns are numeric
grade_scale <- c("A" = 4, "B" = 3, "C" = 2, "D" = 1, "F" = 0)
school_data <- school_data %>% mutate(numeric_grade = grade_scale[overallgrade])

cor_data <- school_data %>%
  select(numeric_grade, peerindex, overallscore2012, overallpercentile2012,
         progressscore2012, performancescore2012, environmentscore2012,
         additionalcredit2012, parentresponserate, teacherresponserate, 
         studentresponserate, safetyscore, communicationscore, 
         engagementscore, expectationsscore) %>%
  mutate(across(everything(), ~as.numeric(.)))  # Convert all columns to numeric

# Drop rows with missing values
cor_data <- na.omit(cor_data)

# Compute correlation matrix
cor_matrix <- cor(cor_data)

# Visualize the correlation matrix
library(pheatmap)
pheatmap(cor_matrix,
         main = "Correlation Heatmap",
         color = colorRampPalette(c("blue", "white", "red"))(50),
         display_numbers = TRUE,
         fontsize_number = 8)


```
```{r}
# Create correlation matrix
cor_data <- school_data %>%
  select(numeric_grade, peerindex, overallscore2012, overallpercentile2012,
         progressscore2012, performancescore2012, environmentscore2012,
         additionalcredit2012, parentresponserate, teacherresponserate, 
         studentresponserate, safetyscore, communicationscore, 
         engagementscore, expectationsscore) %>%
  mutate(across(everything(), ~as.numeric(.)))

cor_data <- na.omit(cor_data)
cor_matrix <- cor(cor_data, use = "complete.obs")


# Convert to long format for Tableau
cor_long <- as.data.frame(as.table(cor_matrix))

# Save as CSV
write.csv(cor_long, "correlation_matrix_long.csv", row.names = FALSE)

```

---

Convert Overall Grades to Numeric Values

```{r}
grade_scale <- c("A" = 4, "B" = 3, "C" = 2, "D" = 1, "F" = 0)

school_data <- school_data %>% mutate(numeric_grade = grade_scale[overallgrade])
```

---

Average Grade by District

```{r}
avg_grade_by_district <- school_data %>%
  group_by(district) %>%
  summarise(average_grade = mean(numeric_grade, na.rm = TRUE)) %>%
  arrange(desc(average_grade))

print(avg_grade_by_district)
```

---

## Average Climate Benchmarks by District

```{r}
district_comparison <- school_data %>%
  group_by(district) %>%
  summarise(
    avg_grade = mean(numeric_grade, na.rm = TRUE),
    avg_safety = mean(safetyscore, na.rm = TRUE),
    avg_communication = mean(communicationscore, na.rm = TRUE),
    avg_engagement = mean(engagementscore, na.rm = TRUE),
    avg_expectations = mean(expectationsscore, na.rm = TRUE)
  ) %>%
  arrange(desc(avg_grade))

print(district_comparison)
```
```{r}
district_comparison <- school_data %>%
  group_by(district) %>%
  summarise(
    avg_grade = mean(numeric_grade, na.rm = TRUE),
    combined_benchmark = mean(
      rowMeans(across(c(safetyscore, communicationscore, engagementscore, expectationsscore)), na.rm = TRUE),
      na.rm = TRUE
    )
  ) %>%
  arrange(desc(combined_benchmark))  # Optional sorting

print(district_comparison)


```
```{r}
correlation <- cor(district_comparison$avg_grade, district_comparison$combined_benchmark, use = "complete.obs")
print(paste("Correlation:", round(correlation, 3)))

ggplot(district_comparison, aes(x = combined_benchmark, y = avg_grade)) +
  geom_point(color = "darkblue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Do Higher Benchmarks Predict Higher Grades?",
    x = "Combined Benchmark Score",
    y = "Average Grade"
  )
model <- lm(avg_grade ~ combined_benchmark, data = district_comparison)
summary(model)



```

---

## Visualize Grades vs Benchmark Scores

```{r}
district_long <- district_comparison %>%
  pivot_longer(cols = starts_with("avg_"), names_to = "metric", values_to = "value")

ggplot(district_long, aes(x = factor(district), y = value, fill = metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Average Grade vs. Benchmark Scores by District",
    x = "District",
    y = "Score",
    fill = "Metric"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

---

## Map Benchmark Score to Letter Grade and Compare

```{r}
# Convert relevant scores to numeric
school_data <- school_data %>%
  mutate(
    progressscore2012 = as.numeric(progressscore2012),
    performancescore2012 = as.numeric(performancescore2012),
    environmentscore2012 = as.numeric(environmentscore2012),
    additionalcredit2012 = as.numeric(additionalcredit2012)
  )

# Create comparison table
score_analysis_by_district <- school_data %>%
  group_by(district) %>%
  summarise(
    avg_numeric_grade = mean(numeric_grade, na.rm = TRUE),
    avg_progress_score = mean(progressscore2012, na.rm = TRUE),
    avg_performance_score = mean(performancescore2012, na.rm = TRUE),
    avg_environment_score = mean(environmentscore2012, na.rm = TRUE),
    avg_additional_credit = mean(additionalcredit2012, na.rm = TRUE)
  ) %>%
  mutate(
    avg_benchmark_score = rowMeans(select(., avg_progress_score, avg_performance_score, avg_environment_score, avg_additional_credit), na.rm = TRUE),
    benchmark_letter = case_when(
      avg_benchmark_score >= 45 ~ "A",
      avg_benchmark_score >= 35 ~ "B",
      avg_benchmark_score >= 25 ~ "C",
      avg_benchmark_score >= 15 ~ "D",
      TRUE ~ "F"
    ),
    actual_letter = names(grade_scale)[round(avg_numeric_grade)]
  )

# Visualize comparison of letter grades
viz_data <- score_analysis_by_district %>%
  select(district, actual_letter, benchmark_letter) %>%
  pivot_longer(cols = c(actual_letter, benchmark_letter), names_to = "grade_type", values_to = "grade")

ggplot(viz_data, aes(x = factor(district), fill = grade)) +
  geom_bar(position = "dodge") +
  facet_wrap(~grade_type) +
  labs(
    title = "Actual vs Benchmark Letter Grades by District",
    x = "District",
    y = "Count",
    fill = "Grade"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

---
```{r}
library(dplyr)

# First, make sure you have a numeric version of the overall grade
grade_scale <- c("A" = 4, "B" = 3, "C" = 2, "D" = 1, "F" = 0)

school_data <- school_data %>%
  mutate(numeric_grade = grade_scale[overallgrade])

# Now group by school_type and get the average
avg_grade_by_school_type <- school_data %>%
  group_by(school_type) %>%
  summarise(
    average_grade = mean(numeric_grade, na.rm = TRUE),
    count = n()
  ) %>%
  arrange(desc(average_grade))

print(avg_grade_by_school_type)


ggplot(avg_grade_by_school_type, aes(x = reorder(school_type, -average_grade), y = average_grade)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Grade by School Type",
    x = "School Type",
    y = "Average Grade (Numeric)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
Is community involvement a key factor
```{r}
cor(school_data$parentresponserate, school_data$engagementscore, use = "complete.obs")
lm(engagementscore ~ parentresponserate + teacherresponserate + school_type, data = school_data)

```
--

Is there correlation between schools hitting or missing certain benchmarks?

```{r}
school_data <- school_data %>%
  mutate(
    hit_engagement = ifelse(engagementscore >= 8, 1, 0),
    hit_expectations = ifelse(expectationsscore >= 8, 1, 0)
  )

# Correlation between hitting two benchmarks
cor(school_data$hit_engagement, school_data$hit_expectations, use = "complete.obs")

```
Does parent or staff involvement affect how students hit benchmarks?
```{r}
summary(lm(engagementscore ~ parentresponserate + teacherresponserate, data = school_data))
```
Which school chnaged there grade the mosst from progress to final
```{r}
# Grade mapping function
grade_to_num <- function(grade) {
  return(switch(grade,
                "A" = 4,
                "B" = 3,
                "C" = 2,
                "D" = 1,
                "F" = 0,
                NA))  # return NA if the grade is NA
}

# Apply the mapping to progress and overall grades
school_data$progress_numeric <- sapply(school_data$progressgrade2012, grade_to_num)
school_data$final_numeric <- sapply(school_data$overallgrade, grade_to_num)

# Calculate grade change
school_data$grade_change <- school_data$final_numeric - school_data$progress_numeric

sorted_data <- school_data[order(-abs(school_data$grade_change), school_data$grade_change), ]

# View top rows with largest changes
head(sorted_data[, c("school_name", "progressgrade2012", "overallgrade", "grade_change")], 20)

```


What factors are most strongly correlated with a school's overall grade?
```{r}

# Apply conversion to overall grade
school_data$overall_numeric <- sapply(school_data$overallgrade, grade_to_num)

# Convert other relevant columns to numeric
numeric_columns <- c("peerindex", "parentresponserate", "teacherresponserate",
                     "safetyscore", "communicationscore", "engagementscore",
                     "expectationsscore", "additionalcredit2012")

# Make sure all are numeric
school_data[numeric_columns] <- lapply(school_data[numeric_columns], function(x) as.numeric(as.character(x)))

# Create a subset for correlation
cor_data <- school_data[, c("overall_numeric", numeric_columns)]

# Remove rows with NA
cor_data <- na.omit(cor_data)

# Compute correlation matrix
cor_matrix <- cor(cor_data, use = "complete.obs")

# View correlation with overall grade
cor_with_overall <- cor_matrix["overall_numeric", ]
cor_with_overall <- sort(cor_with_overall[-1], decreasing = TRUE)  # exclude self-correlation

# Print
print("Correlation of features with Overall Grade (Numeric):")
print(round(cor_with_overall, 3))

```

Do schools with higher parent/teacher response rates tend to have better overall grades?
```{r}
school_data$overall_numeric <- sapply(school_data$overallgrade, grade_to_num)

# Convert response rates to numeric
school_data$parentresponserate <- as.numeric(school_data$parentresponserate)
school_data$teacherresponserate <- as.numeric(school_data$teacherresponserate)

# Remove rows with missing values
sub_data <- na.omit(school_data[, c("overall_numeric", "parentresponserate", "teacherresponserate")])

# Compute correlation
cor_parent <- cor(sub_data$overall_numeric, sub_data$parentresponserate)
cor_teacher <- cor(sub_data$overall_numeric, sub_data$teacherresponserate)

# Print results
cat("Correlation between overall grade and parent response rate:", round(cor_parent, 3), "\n")
cat("Correlation between overall grade and teacher response rate:", round(cor_teacher, 3), "\n")

library(ggplot2)

# Scatter plot with trend line for parent response
ggplot(sub_data, aes(x = parentresponserate, y = overall_numeric)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Parent Response Rate vs Overall Grade",
       x = "Parent Response Rate", y = "Overall Grade (Numeric)")

```
Excluding elemenatry schools
```{r}
# Grade mapping
school_data$overall_numeric <- sapply(school_data$overallgrade, grade_to_num)

# Convert response rates to numeric
school_data$parentresponserate <- as.numeric(school_data$parentresponserate)
school_data$teacherresponserate <- as.numeric(school_data$teacherresponserate)

# 🔍 Exclude elementary schools
non_elem_data <- subset(school_data, !grepl("Elementary", school_type, ignore.case = TRUE))

# Remove rows with NA
sub_data <- na.omit(non_elem_data[, c("overall_numeric", "parentresponserate", "teacherresponserate")])

# Correlation
cor_parent <- cor(sub_data$overall_numeric, sub_data$parentresponserate)
cor_teacher <- cor(sub_data$overall_numeric, sub_data$teacherresponserate)

# Print results
cat("Correlation (excluding Elementary) - Parent:", round(cor_parent, 3), "\n")
cat("Correlation (excluding Elementary) - Teacher:", round(cor_teacher, 3), "\n")

# Plot for Parent Response
library(ggplot2)
ggplot(sub_data, aes(x = parentresponserate, y = overall_numeric)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Parent Response Rate vs Overall Grade (Non-Elementary)",
       x = "Parent Response Rate", y = "Overall Grade (Numeric)")

```






Which school Types tend to perform better overall
```{r}

school_data$overall_numeric <- sapply(school_data$overallgrade, grade_to_num)

# Group by school type and compute average grade
avg_by_type <- aggregate(overall_numeric ~ school_type, school_data, mean, na.rm = TRUE)

# Sort descending
avg_by_type <- avg_by_type[order(-avg_by_type$overall_numeric), ]

print(avg_by_type)

```
Do districts who have higher response rates are the ones with the higher grades?
```{r}

# Convert response rates to numeric
school_data$parentresponserate <- as.numeric(school_data$parentresponserate)
school_data$teacherresponserate <- as.numeric(school_data$teacherresponserate)

# Group by district and compute averages
library(dplyr)

district_summary <- school_data %>%
  filter(!is.na(overall_numeric), !is.na(parentresponserate), !is.na(teacherresponserate)) %>%
  group_by(district) %>%
  summarise(
    avg_grade = mean(overall_numeric, na.rm = TRUE),
    avg_parent_response = mean(parentresponserate, na.rm = TRUE),
    avg_teacher_response = mean(teacherresponserate, na.rm = TRUE),
    n_schools = n()
  )

# Correlation between avg response rates and avg grade
cor_parent <- cor(district_summary$avg_grade, district_summary$avg_parent_response)
cor_teacher <- cor(district_summary$avg_grade, district_summary$avg_teacher_response)

cat("Correlation between average parent response rate and average grade:", round(cor_parent, 3), "\n")
cat("Correlation between average teacher response rate and average grade:", round(cor_teacher, 3), "\n")

library(ggplot2)

ggplot(district_summary, aes(x = avg_parent_response, y = avg_grade)) +
  geom_point(color = "blue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue") +
  labs(title = "Average Parent Response Rate vs Average Grade by District",
       x = "Average Parent Response Rate", y = "Average Overall Grade")

# Repeat with teacher
ggplot(district_summary, aes(x = avg_teacher_response, y = avg_grade)) +
  geom_point(color = "orange", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  labs(title = "Average Teacher Response Rate vs Average Grade by District",
       x = "Average Teacher Response Rate", y = "Average Overall Grade")


```
Does improvement in a specific benchmark (e.g., safety) predict an increase in grades better than others?
```{r}
lm(numeric_grade ~ safetyscore + communicationscore + engagementscore + expectationsscore, data = school_data)

```

Clustering
```{r}
library(dplyr)
library(ggplot2)

# Select and scale relevant climate benchmark features
cluster_data <- school_data %>%
  select(safetyscore, communicationscore, engagementscore, expectationsscore) %>%
  na.omit()  # remove incomplete rows

scaled_cluster_data <- scale(cluster_data)

# Elbow Method to determine optimal clusters
wss <- vector()
for (k in 1:10) {
  kmeans_result <- kmeans(scaled_cluster_data, centers = k, nstart = 25)
  wss[k] <- kmeans_result$tot.withinss
}
plot(1:10, wss, type = "b", pch = 19,
     xlab = "Number of Clusters (k)", ylab = "Total Within-Cluster Sum of Squares",
     main = "Elbow Method for Optimal k")

# Apply k-means clustering with chosen k
set.seed(42)
kmeans_result <- kmeans(scaled_cluster_data, centers = 3, nstart = 25)

# Attach original cluster labels back to full dataset
school_data$cluster <- NA
school_data$cluster[as.numeric(rownames(cluster_data))] <- kmeans_result$cluster

# ---- REORDER CLUSTERS BY AVERAGE GRADE ----
# Step 1: Summarize current clusters by avg grade
cluster_summary <- school_data %>%
  filter(!is.na(cluster)) %>%
  group_by(cluster) %>%
  summarise(avg_grade = mean(overall_numeric, na.rm = TRUE)) %>%
  arrange(avg_grade)

# Step 2: Create mapping from old to new cluster labels
cluster_order_map <- setNames(1:nrow(cluster_summary), cluster_summary$cluster)

# Step 3: Apply new labels to school_data
school_data$cluster_reordered <- cluster_order_map[as.character(school_data$cluster)]

# Step 4: Resummarize by reordered clusters
reordered_summary <- school_data %>%
  filter(!is.na(cluster_reordered)) %>%
  group_by(cluster_reordered) %>%
  summarise(
    avg_grade = mean(overall_numeric, na.rm = TRUE),
    avg_safety = mean(safetyscore, na.rm = TRUE),
    avg_communication = mean(communicationscore, na.rm = TRUE),
    avg_engagement = mean(engagementscore, na.rm = TRUE),
    avg_expectations = mean(expectationsscore, na.rm = TRUE),
    n_schools = n()
  )

print(reordered_summary)

# Visualize reordered clusters using PCA
pca <- prcomp(scaled_cluster_data)
pca_df <- as.data.frame(pca$x[, 1:2])  # first two principal components
pca_df$cluster <- factor(school_data$cluster_reordered[as.numeric(rownames(cluster_data))])

ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.6, size = 3) +
  labs(title = "School Clusters Based on Climate Benchmarks",
       x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()


```

```{r}
library(ggplot2)
library(tidyr)

# Create your cluster summary data manually or from previous code
cluster_summary <- data.frame(
  cluster = factor(1:4),
  avg_grade = c(3.222222, 1.989071, 2.681495, 2.996310),
  avg_safety = c(8.761765, 6.653070, 7.449340, 8.126876),
  avg_communication = c(8.245425, 6.203070, 6.891254, 7.528272),
  avg_engagement = c(8.487908, 6.494298, 7.199175, 7.812042),
  avg_expectations = c(8.692157, 6.838158, 7.545050, 8.107155)
)

# Convert to long format for ggplot
cluster_long <- cluster_summary %>%
  pivot_longer(cols = -cluster, names_to = "Metric", values_to = "Value")

# Bar plot
ggplot(cluster_long, aes(x = Metric, y = Value, fill = cluster)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Cluster Comparison of Grades and Climate Benchmarks",
       x = "Metric", y = "Average Score") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")

```
Which Benchamrk is hardest to hit
```{r}
# Load necessary libraries
library(dplyr)

# Select only the benchmark columns
benchmark_data <- school_data %>%
  select(safetyscore, communicationscore, engagementscore, expectationsscore) %>%
  mutate(across(everything(), as.numeric)) %>%
  na.omit()

# Calculate the average score for each benchmark
benchmark_means <- colMeans(benchmark_data)

# Print the means
print(benchmark_means)

# Identify the hardest benchmark (lowest average)
hardest_benchmark <- names(benchmark_means)[which.min(benchmark_means)]
cat("The hardest benchmark to hit is:", hardest_benchmark, "with a mean score of", round(min(benchmark_means), 2), "\n")


```
regression analysis
```{r}
# Convert overall grade to numeric if not already done
grade_scale <- c("A" = 4, "B" = 3, "C" = 2, "D" = 1, "F" = 0)
school_data$numeric_grade <- grade_scale[school_data$overallgrade]

# Fit linear regression model
model <- lm(numeric_grade ~ safetyscore + communicationscore + engagementscore +
              expectationsscore + parentresponserate + teacherresponserate +
              peerindex, data = school_data)

# View model summary
summary(model)

```


Which school type has the highest average overall grade?
```{r}
school_data %>%
  group_by(school_type) %>%
  summarise(avg_grade = mean(overall_numeric, na.rm = TRUE)) %>%
  arrange(avg_grade)

```

YABC = Young adult borough centers: afternoon and evening programs designed to meet the educational needs of high school students who are behind in credit or have adult responsibilities that make it hard to attend school during the day


```{r}
# Select only the benchmark columns
benchmark_data <- school_data %>%
  select(safetyscore, communicationscore, engagementscore, expectationsscore) %>%
  mutate(across(everything(), ~as.numeric(.))) %>%
  na.omit()

# Calculate correlation matrix
benchmark_cor <- cor(benchmark_data)

# View as table
print(round(benchmark_cor, 2))

# Optional: Visualize as heatmap
library(pheatmap)

pheatmap(benchmark_cor,
         main = "Correlation Between School Climate Benchmarks",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(0, 1, length.out = 101),  # Force color scale from 0 to 1
         display_numbers = TRUE,
         fontsize_number = 10,
         border_color = NA)


```


```{r}
# Assuming `school_data` is your dataframe and has columns: district, parentresponserate, numeric_grade

# Convert overall grade to numeric if not already done
grade_scale <- c("A" = 4, "B" = 3, "C" = 2, "D" = 1, "F" = 0)
school_data$numeric_grade <- grade_scale[school_data$overallgrade]

# Group by district and compute mean values
district_averages <- school_data %>%
  group_by(district) %>%
  summarise(
    avg_grade = mean(numeric_grade, na.rm = TRUE),
    avg_parent_response = mean(parentresponserate, na.rm = TRUE)
  )

# Calculate correlation
correlation <- cor(district_averages$avg_grade, district_averages$avg_parent_response, use = "complete.obs")

# Print result
print(correlation)

```

```{r}
# Step 1: Convert grades to numeric if needed
grade_scale <- c("A" = 4, "B" = 3, "C" = 2, "D" = 1, "F" = 0)
school_data$numeric_grade <- grade_scale[school_data$overallgrade]

# Step 2: Define threshold and flag benchmarks hit per school (e.g., score ≥ 8)
school_data <- school_data %>%
  mutate(
    hit_safety = safetyscore >= 8,
    hit_communication = communicationscore >= 8,
    hit_engagement = engagementscore >= 8,
    hit_expectations = expectationsscore >= 8,
    total_hits = hit_safety + hit_communication + hit_engagement + hit_expectations
  )

# Step 3: View correlation (optional)
cor(school_data$total_hits, school_data$numeric_grade, use = "complete.obs")

# Step 4: Create a summary table per school
library(ggplot2)

ggplot(school_data, aes(x = factor(total_hits), y = numeric_grade)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs(
    title = "Grades by Number of Benchmarks Hit per School",
    x = "Number of Benchmarks Hit (out of 4)",
    y = "Numeric Grade (0 = F, 4 = A)"
  ) +
  theme_minimal()

```

```{r}
# Select just the benchmark columns
benchmarks <- school_data %>%
  select(safetyscore, communicationscore, engagementscore, expectationsscore) %>%
  na.omit()

# Compute correlation matrix
cor_matrix <- round(cor(benchmarks), 2)

# Convert to long format for Tableau
library(reshape2)
cor_long <- melt(cor_matrix)
colnames(cor_long) <- c("Variable1", "Variable2", "Correlation")

# Save for Tableau
write.csv(cor_long, "Benchmark_Correlation_Heatmap.csv", row.names = FALSE)

```

